# Reproducible Research:  Course Project 1

## Loading and preprocessing the data
```{r loaddata}
activity <- read.csv("activity.csv")
```

## What is mean total number of steps taken per day?
```{r}
library(ggplot2)
#### total number of steps per day
total.steps <- tapply(activity$steps, activity$date, FUN=sum, na.rm=TRUE)
#### plot of the steps taken per day
qplot(total.steps, binwidth=1000, xlab="total number of steps taken each day")
#### mean and median of the total number of steps taken per day
mean(total.steps, na.rm=TRUE)
median(total.steps, na.rm=TRUE)
```

## What is the average daily activity pattern?
```{r}
library(ggplot2)
#### timeseries plot of daily activity pattern
averages <- aggregate(x=list(steps=activity$steps), by=list(interval=activity$interval),
                      FUN=mean, na.rm=TRUE)
ggplot(data=averages, aes(x=interval, y=steps)) +
  geom_line() +
  xlab("5-minute interval") +
  ylab("average number of steps taken")
```
On average across all the days in the dataset, the 5-minute interval contains
```{r}
averages[which.max(averages$steps),]

```

## Imputing missing values
There are many days/intervals where there are missing values (coded as `NA`). The presence of missing days may introduce bias into some calculations or summaries of the data.
```{r}
#### Total number of rows with missing data
print(paste("The total number of rows with NA is: ",sum(is.na(activity$steps)))) 
```

All of the missing values are filled in with mean value for that 5-minute
interval.
```{r}
#### Strategy for filling in all of the missing values in the dataset: Replace each missing value with the mean value of its 5-minute interval
fill.value <- function(steps, interval) {
  filled <- NA
  if (!is.na(steps))
    filled <- c(steps)
  else
    filled <- (averages[averages$interval==interval, "steps"])
  return(filled)
}
filled.activity <- activity
filled.activity$steps <- mapply(fill.value, filled.activity$steps, filled.activity$interval)
  
# Histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day.   
total.steps <- tapply(filled.activity$steps, filled.activity$date, FUN=sum)
qplot(total.steps, binwidth=1000, xlab="total number of steps taken each day")
mean(total.steps)
median(total.steps)  

```

Mean and median values are higher after imputing missing data. The reason is that in the original data, there are some days with `steps` values `NA` for  any `interval`. The total number of steps taken in such days are set to 0s by default. However, after replacing missing `steps` values with the mean `steps` of associated `interval` value, these 0 values are removed from the histogram of total number of steps taken each day.

## Are there differences in activity patterns between weekdays and weekends?
First, let's find the day of the week for each measurement in the dataset. In
this part, we use the dataset with the filled-in values.

```{r}
weekday.or.weekend <- function(date) {
  day <- weekdays(date)
  if (day %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"))
    return("weekday")
  else if (day %in% c("Saturday", "Sunday"))
    return("weekend")
  else
    stop("invalid date")
}
filled.activity$date <- as.Date(filled.activity$date)
filled.activity$day <- sapply(filled.activity$date, FUN=weekday.or.weekend)
```
Now, let's make a panel plot containing plots of average number of steps taken
on weekdays and weekends.
```{r}
averages <- aggregate(steps ~ interval + day, data=filled.activity, mean)
ggplot(averages, aes(interval, steps)) + geom_line() + facet_grid(day ~ .) +
  xlab("5-minute interval") + ylab("Number of steps")
```


